
\documentclass{article}

\usepackage{amsmath}

\newcommand{\A}{\mathcal A}
\newcommand{\C}{\mathcal C}

\begin{document}

In this text, I summarize and discuss some details about LocARNA's scoring.

\section{MEA scoring}
\label{sec:meascoring}

The new mea score is a weighted sum of three components

\begin{enumerate}
\item alignment accuracy
\item structural accuracy
\item consensus accuracy
\end{enumerate}

\paragraph{ad 1.)} Alignment accuracy adds for each alignment edge
\begin{displaymath}
  Pr[i\sim j|A,B],
\end{displaymath}
which is the probability to align the sequence positions $i$ and $j$ in a
sequence alignment of $A$ and $B$ (with affine gap penalty).  We compute these
probabilities using a local partition variant of Gotoh's algorithm.

\paragraph{ad 2.)} Structural accuracy has two components and contributes for each
sequence. First, it adds for each arc-match/consensus basepair the
corresponding basepair probability. Second, it adds for each non-structured
edge the probability that the position is unpaired.

\paragraph{ad 3.)} Consensus accuracy adds for each arc match/consensus basepair the term
\begin{displaymath}
  Pr[i\sim k|AB] \cdot Pr[j\sim l|AB] \cdot \frac{f'_{ijkl}}{f_{ij}f_{kl}}.
\end{displaymath}

\subsection{Weights and total score}

For relating the components to each other we introduce weights $\alpha$,
$\beta$, $\gamma$.

The total score is then
\begin{displaymath}
  \sum_{ij\in\A} Pr[i\sim j|AB] + \alpha \sum_{i\not\in\C} ( \bar p^A_{\pi^1(i)} +  \bar
  p^B_{\pi^2(i)} ) + \beta \sum_{(ij)\in\C} (p^A_{\pi^1(ij)} + p^B_{\pi^2(ij)})
  +  \gamma \sum_{(ij)\in C} Pr[i\sim k|AB] \cdot Pr[j\sim l|AB] \cdot \frac{f'_{ijkl}}{f_{ij}f_{kl}}.
\end{displaymath}
{\bf die obige Formel stimmt mit den Projektionen nicht, man sollte dafuer mal eine bessere
  Schreibweise finden.}

The score can be divided into a contribution per basematch/alignment edge
and a contribution per arcmatch/structural alignment edge/consensus basepair.



\end{document}



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
